---
title: "COVID-19 Data"
output: html_document
date: "2024-10-01"
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(tidyverse)
library(dplyr)
```

## Data Import

I imported the data into the new variable url_in and file_names.

```{r URLS, include=TRUE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names<-
  c("time_series_covid19_confirmed_global.csv",
    
    "time_series_covid19_deaths_global.csv",
    "time_series_covid19_confirmed_US.csv",
    "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)
              
```

Let's read in the data and see what we have.

```{r import_data, message=FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```

After looking at global_cases and global_deaths, I would like to tidy those datasets and put each variable (date, cases, deaths) in their own column.
Also, I don't need Lat and Long for the analysis I am planning, so I will get rid of those and rename Region and State to be more R friendly.

```{r tidy_global_data, include=TRUE}
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat, Long))
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))
 global <- global %>% filter(cases > 0)
US_cases %>%
pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases")
US_cases <- US_cases %>%
pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases") %>%
select(Admin2:cases) %>%
mutate(date = mdy(date)) %>%
select(-c(Lat, Long_))
US_cases
US_deaths <- US_deaths %>%
pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths") %>%
select(Admin2:deaths) %>%
mutate(date = mdy(date)) %>%
select(-c(Lat, Long_))

US <- US_cases %>%
full_join(US_deaths)

global <- global %>%
unite("Combined_Key", c(Province_State, Country_Region), sep = ", ",
na.rm = TRUE,
remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

```

## Visualization

Now I will create ways to plot my tidy data in a user friendly way that is more understandable to the untrained eye.

```{r visuals, include=TRUE}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_totals <- US_by_state %>%
group_by(Country_Region, date) %>%
summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
mutate(deaths_per_mill = deaths *1000000 / Population) %>%
select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
ungroup()

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases))+
  geom_line(aes(color = "cases"))+
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths"))+
  geom_point(aes(y = deaths, color = "deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))+
  labs(title = "COVID19 in US", y = NULL)

state <- "New York" 
US_by_state %>%
filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases))+
  geom_line(aes(color = "cases"))+
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths"))+
  geom_point(aes(y = deaths, color = "deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))+
  labs(title = str_c("COVID19 in ", state), y = NULL)

max(US_totals$date)
max(US_totals$deaths)

US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
tail(US_totals)

tail(US_totals %>% select(new_cases, new_deaths, everything()))

US_totals %>%
  ggplot(aes(x = date, y = new_cases))+
  geom_line(aes(color = "new_cases"))+
  geom_point(aes(color = "new_cases"))+
  geom_line(aes(y = new_deaths, color = "new_deaths"))+
  geom_point(aes(y = new_deaths, color = "new_deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))+
  labs(title = "COVID19 in US", y = NULL)

state <- "New York" 
US_by_state %>%
filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases))+
  geom_line(aes(color = "new_cases"))+
  geom_point(aes(color = "new_cases"))+
  geom_line(aes(y = new_deaths, color = "new_deaths"))+
  geom_point(aes(y = new_deaths, color = "new_deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))+
  labs(title = str_c("COVID19 in ", state), y = NULL)
```

## Models & Analysis

Now we can calculate some more complex and instructive values from our raw data. 


```{r models and analysis, include=TRUE}
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases), Population = max(Population), cases_per_thou = 1000*cases / Population, deaths_per_thou = 1000* deaths / Population) %>%
  filter(cases > 0, Population > 0)

US_state_totals
```


## Conclusion & Sources of Bias

Covid-19 was a emotionally heated epidemic. Many people lost loved ones. With that in mind, it seems natural that some emotional bias might make it's way into the analysis of this data. For example, one might highlight outliers that exaggerate the deaths or cases of the entire stud, rather than focus on the more accurate data sets. 

In conclusion, we have found that Covid-19 has resulted in the deaths on average of 3 people per thousand within the US. There have been considerable correlation between cases of Covid-19 and resulting deaths. Furthermore, as the visuals have shown, there have been periods of latency within the pandemic and periods of spiked cases in a sort of cyclical manner. 